---
title: "KeyMethods used in the baseline paper"
output: html_document
---
This is a personal document to demonstrate the key methods used in the paper with code snippets.  
<http://nature.com/articles/s41591-020-0769-8>

## Predictive performace using ROC (pROC R package) and AUC

ROC step:
1. Firstly fit a logistic regression (e.g. Response ~ CD38hi) and use the built model (fitted values) to compute True Positve Rate (TPR) and False Positive Rate (FPR) at various thresholds (from 0 to 1)  
-> Remember **Sensitivity**, **Specificity**, **False Positive Rate** can be calculated as  
(From a binary classification confusion matrix):  

Sensitivity = (True Positive) / (True Positive + False Negative (Type 2 Error)) = True Positive Rate (TPR)  
Specificity = (True Negative) / (True Negative + False Positive (Type 1 Error)) = True Negative Rate (TNR)  
False Positive Rate = 1 - Specificity  
  
-> The resulting plot (TPR vs FPR) represents the ROC plot (AUC tells how good the model is)

2. To get a p-value, a random permutation test is used - first permute the 'CD38hi' feature. So the original relationship with the 'Response' no longer exists.  

Then, ROC is constructed for each permutation (e.g. 1000 permutations) and AUCs are stored.

```{r setup, include=FALSE}

## df = A dataframe consist of:
##    1. timepoints (-7, 0, 70 baseline timepoints)
##    2. CD38hi %
##    3. Response (high, low)
##    4. Subjects

roc.df = data.frame()
df.text = data.frame()
N_perm = 1000
for (tp in c(0,-7,70)) {
  r = roc(Response~CD38hi, data=dplyr::filter(df,time==tp), direction="<", quiet = T)
  set.seed(123)
  r.null = vector("numeric", N_perm)
  for(i in 1:N_perm) {
    # iperm = sample(nrow(df))
    r.null[i] = roc(Response~sample(CD38hi), data=dplyr::filter(df,time==tp), direction="<", quiet = T)$auc
  }
  r.p = sum(r.null > r$auc) / N_perm # This calculates the p-value from the distribution derived from the random permutation (n = 1000)
  r.df = data.frame(day=paste0("day ",tp), Specificity = r$specificities, Sensitivity=r$sensitivities) %>%
    arrange(Sensitivity)
  roc.df = rbind(roc.df, r.df)
  
  df.text = rbind(df.text, 
                  data.frame(day=paste0("day ",tp), label=sprintf("AUC = %.2f\np = %.2g",r$auc,r.p)))
}
df.text$x = rep(0.25,3)
df.text$y = rep(0.1,3)
df.text = df.text %>% 
  mutate(day = factor(day, levels=levels(roc.df$day)))

## roc.df holds the specificties and sensitivities for all 3 baseline timepoints (0, 7, 70)

```

## ANOVA to test temporal stability for each genes

1. For each gene and all the subjects (-7, 0 70 timepoints)  
2. ANOVA test: Within the group variance (SSW) + between the group variance (SSB) = total variance.  
3. Ratio - F - statistic (SSB/SSW) and p-values are computed.  
4. Finally, p-values are adjusted.  

```{r}

# stability calculation
ss = matrix(NA,nrow(dat), 2, dimnames=list(rownames(dat),c("ISV","WSV"))) # Intra-subject variation (ISV) & Within-subject variation (WSV)
pv = rep(NA,nrow(dat))
names(pv) = rownames(dat)
for (i in 1:nrow(dat)) {
  tmp = info %>% dplyr::select(subject) %>% mutate(value=dat[i, ])
  fit = aov(value~subject, data=tmp)
  ss[i,] = summary(fit)[[1]]["Sum Sq"][[1]]
  pv[i] = summary(fit)[[1]]["Pr(>F)"][[1]][1]
}

pv.bh = p.adjust(pv, "BH")
pv.bonf = p.adjust(pv, "bonferroni")
ssn = ss / rowSums(ss)

fn.ssn = file.path(PROJECT_DIR, "generated_data", "CHI", "CHI_genes_stability.txt")

ssn %>% as.data.frame() %>% tibble::rownames_to_column("gene") %>% 
  mutate(pv = pv, pv.BH=pv.bh, pv.Bonf=pv.bonf) %>% 
  fwrite(fn.ssn, sep="\t", col.names=T, row.names=F, quote=F)

```

## Correlation of temporal stable genes and CD38hi B cells

1. To get the robust correlation, correlations are calculated for multiple iterations.    
2. Per iteration, one unique pair of samples is removed and correlation (Speaman's rank correlation) is calculated.  
3. Iteration is over when all the unique pairs have been removed.  
4. Parameters that we are interested are calculated from the correlation values (e.g. mean, sd, correlation coefficient (mean/sd))

In this paper, this was done using:  
  a) temporal stable genes (ISV > 0.75)
  b) all genes + repeat 500 times (for random signatures)
  
** select top 'n' genes based on the correlation coefficients and plot ROCs (response ~ avg z-score of the top n genes) to get AUC 
using a) temporal stable genes and b) all genes (for all 500 random iterations) to generate a null distribution

```{r}

# dat = gene-expression matrix that only contains temporal-stable genes

ns = 2
cmb = combn(1:nrow(info), ns) # cmb is all the combinations of different pairs
# cc.rob = matrix(NA,nrow=nrow(edat.gene),ncol=ncol(cmb))

cc.rob = matrix(NA,nrow=sum(gidx),ncol=ncol(cmb))

for (ic in 1:ncol(cmb)) {
  cat(ic," ")
  ridx = cmb[,ic]
  cc.rob[,ic] = cor(t(dat[gidx,-ridx]), info$CD38hi[-ridx], method="spearman", use="pairwise.complete.obs")
  # correlation = (~720 genes x 22 samples , 22 samples CD38Hi values) -> 1D array (correlations for each gene)
}
rownames(cc.rob) = rownames(dat)[gidx]
# columns of the cc.rob represent all the correlation structures calculated above

cc.rob.rank = apply(-cc.rob,2,rank)
# -cc.rob so rank() function gives the most positive correlated value -> 1

ntop = 20
cc.rob.ntop = rowSums(cc.rob.rank <= ntop)
# How many correlations structures have genes that are ranked in the top 20?

cc.rob.mean = apply(cc.rob,1,mean) # gene-wise mean correlation
cc.rob.median = apply(cc.rob,1,median) # gene-wise median correlation
cc.rob.sd = apply(cc.rob,1,sd) # gene-wise SD correlation
cc.rob.1cv = cc.rob.mean / cc.rob.sd # gene-wise CV correlation
cc.rob.ntop.rank = rank(cc.rob.ntop)

cc.rob.1cv.ord = order(cc.rob.1cv,decreasing = T)
cc.rob.1cv.sort = sort(cc.rob.1cv,decreasing = T)
head(cc.rob.1cv.sort, 10)

# calculate AUC for each gene
auc.one=matrix(nrow=nrow(dat.ts), ncol=3)
for (k in 1:nrow(dat.ts)) {
  X = dat.ts[k,]
  Y = info$Response
  auc.one[k, c(2,1,3)] = ci.auc(Y,X,direction="<", quiet=T)
}
rownames(auc.one) = rownames(dat.ts)
colnames(auc.one) = c("auc.gene", "auc.ci95.min", "auc.ci95.max")

# output
df.out = data.frame(cor.ntop20 = cc.rob.ntop / ncol(cmb) *100, 
                    cor.mean = cc.rob.mean, 
                    cor.sd = cc.rob.sd, 
                    cor.median = cc.rob.median,
                    cor.mean.sd.ratio = cc.rob.1cv, # correlation coefficient
                    as.data.frame(auc.one)) %>%
  tibble::rownames_to_column("gene")

```
